Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit

### 2023-07-03 Manuscript/Manuscript.Rmd
Ms edits and bib updates

Nitric oxide (NO) is one such modulator with unique physical and signalling properties.
This free radical synthesized from L-arginine by nitric oxide synthase (NOS) is short-lived and can diffuse across biological membranes [@Cudeiro1999; @thomas2015].
Canonical NO signalling involves the Ca^2+^/calmodulin-dependent activation of NOS, NO production and diffusion, and the NO-dependent activation of soluble guanylate cyclases (sGC) leading to cGMP production [@Bredt1990; @Hoelscher1997].
Given that NOS activation is Ca^2+^ dependent and NOS shows neuron-type-specific expression [@Gibbs1998a; @Wildemann1999; @Aso2019; @Mobley2022], NO action can lead to the activity-dependent modulation of neural circuits at specific sites [@Wang2007; @Vielma2014; @jacoby2018; @Aso2019] .

In the vertebrate retina, NOS is expressed in amacrine, ganglion and other cells and the actions of NO can be diverse [@Cudeiro1999; @Wang2007; @jacoby2018].
For example, defective NO signalling in NOS knockouts leads to a decreased sensitivity of retinal ganglion cells to light stimulation [@Wang2007].
In the retina, NO signalling can also involve pathways other than canonical sGC signalling [@Tooker2013; @Wei2012; @jacoby2018].
Due to the complex expression of NOS in vertebrates and the diversity of its functions, it has been challenging to link the neurophysiological effects of NO to signalling mechanisms and behaviour change.

@@

```{r echo=FALSE,fig.cap="**Figure 2. NO produced by UV/violet stimulation to cPRCs.** (**A**) DAF-FM fluorescence in the region of the neurosecretory neuropil. The white line indicates the outline of the larva. The dashed line corresponds to the area where fluorescence was quantified. Circles indicate the location of cPRC and control stimulation. cPRCs are marked by thin lines. (**B**) Changes in DAF-FM fluorescence before and during 405 nm light stimulation. (**C**) Changes in DAF-FM fluorescence over time during 405 nm stimulation of the cPRCs (green) or a control area (ctr: gray). The purple box indicates the duration of 405 nm stimulation. Individual traces normalized (ΔF/F0) are shown as thin lines. Thick lines show the mean value with 0.95 confidence intervals. N = 9 larvae for control and 11 for cPRC stimulation."}

@@

To characterise the expression of NIT-GCs, we used previously published spatially mapped single-cell transcriptome data [@Achim2015; @williams2017].
Among the 15 NIT-GCs, two showed high and specific expression in the cPRCs and one was expressed in the INNOS cells (Figure 4---figure supplement 2).
In the single-cell data, we could identify the cPRCs by the specific expression of *c-opsin1* and the *pedal-peptide2 neuropeptide precursor* (*MLD proneuropeptide*), previously described cPRC markers [@Arendt2004; @williams2017] (Figure 4---figure supplement 3A).
The INNOS cells were identified by NOS expression and spatial mapping in the brain [@Achim2015].
We decided to focus on two NIT-GCs expressed in the cPRCs and with a full-length sequence, *NIT-GC1* and *NIT-GC2*.
To confirm the single-cell data, we first carried out in situ hybridisation chain reaction (HCR) with probes for *NIT-GC1* and *NIT-GC2* mRNA. Both genes were specifically expression in the four cPRCs, as confirmed by co-labeling with an acetylated α-tubulin antibody and with an HCR probe against *pedal peptide 2/MLD proneuropeptide* (Figure 4C, D and Figure 4---figure supplement 3A-C).
To analyse the subcellular localisation of NIT-GC1 and NIT-GC2 at the protein level, we raised and affinity-purified polyclonal antibodies against a specific peptide sequence from both proteins.
In immunostainings, we found that NIT-GC1 was localise to the region corresponding to the axonal projections of the cPRCs in the anterior neurosecretory plexus (Figure 4E).
Co-immunostaining with the rabbit NIT-GC1 antibody and a custom rat antibody raised against *Platynereis* NOS revealed the localisation of both proteins in dots in the neurosecretory neuropil (Figure 4---figure supplement 3D).
In contrast, NIT-GC2 specifically labelled the ramified sensory cilia of the cPRCs (Figure 4F).
These different subcellular localisations suggest that the two NIT-GCs are involved in different intracellular signalling processes in the ciliary and axonal regions of the cPRCs.

@@

We first quantified the responses of the INNOS and INRGWa interneurons during 405 nm stimulation of the cPRCs.
In both wild type and *NOS*-mutant larvae, INNOS cells showed an increase in Ca^2+^ during stimulation (Figure 5D).
In contrast, the INNOS response was flat or slightly negative in NIT-GC2 morphant larvae (Figure 5E) revealing an essential role for NIT-GC2-mediated cPRC suppression in INNOS activation.
INRGWa cells were initially inhibited during cPRC stimulation, followed by a delayed activation paralleling the second activation phase of cPRCs.
This late INRGWa response was lacking in *NOS*-mutants (Figure 5F).
In NIT-GC2 morphants, INRGWa cells showed a transient increase in Ca^2+^ that decayed after light off and a delayed activation was not present (Figure 5G).

Next, we imaged Ca^2+^ signals from Ser-h1 and MC neurons in wild type and *NOS* mutant larvae.
Ser-h1 cells showed an activation profile that correlated with cPRC activity, including a reduction in Ca^2+^ during stimulation followed by rebound, a response that was defective in *NOS* mutants (Figure 5I).
MC cells showed sustained activation, including a late-phase that was lacking in *NOS* mutants (Figure 5I).
These data suggest that during 405 nm stimulation the Ser-h1 cells are inhibited and MC cells are activated, and this regulation is NO-dependent.
This pattern is expected to inhibit ciliary activity in the prototroch but not in the other ciliary bands (with no Ser-h1 or MC synapses), triggering NO-dependent downward swimming.

@@

To further analyse the dynamics of responses to UV light and formally describe cPRC phototransduction, synaptic connections and NO retrograde signalling, we developed a mixed cellular-circuit-level mathematical model.
We used our Ca^2+^ imagining data of cPRC, INNOS and INRGWa cells collected in wild type, *NOS* knockout, and *NIT-GC2* morphant larvae to formulate assumptions that are the basis of the proposed model.

We model a direct c-opsin1-dependent response to UV leading to the initial rise in Ca^2+^ levels.
The mechanism of this signal is not known but may include Gα and Gβ signalling, CNGAα channel activation or other pathways.
We further assume that in cPRC cells, Ca^2+^ levels increase with cGMP.
We model a decrease in cGMP and Ca^2+^ levels due to a NIT-GC2-dependent pathway (Figure 6A).
The effect of NO in cPRC cells is captured by another term that describes a NIT-GC1 and NO-dependent increase in cGMP.
Based on the synaptic connectome, we infer a feedforward coupling between cPRC cells and INNOS and INRGWa cells.
The proposed synaptic coupling is UV-dependent, decays linearly and is suppressed in NIT-GC2 morphants (Figure 6A).

In the INNOS cells, we assume that excitatory synaptic input from cPRC (or rebound from tonic inhibition) leads to a rise in Ca^2+^ leading to NO production via NOS.
This input is NIT-GC2-dependent.
In the INRGWa cells, we model a decrease in Ca^2+^ based on an inhibitory UV-dependent synaptic signal from the cPRCs.
We further assume the existence of direct excitatory coupling between cPRC Ca^2+^ levels and INRGWa Ca^2+^ to account for the late effects of cPRC on INRGWa (Figure 6A).
The UV-dependent and Ca^2+^-dependent signals may be mediated by different transmitters released by cPRCs during distinct phases of their activation cycle.
In addition, we assume a direct inhibitory coupling between INNOS Ca^2+^ and INRGWa Ca^2+^.
Finally, in all variables, we assumed a linear decay and constant production to set a steady state (Figure 6A).
Since the aim of the model is to capture the normalised fluorescence data, the model is nondimensionalised.
UV stimulation is modelled as a square pulse.
To find model parameters producing an output fitting the experimental data we employed a global optimisation method known as a genetic algorithm (GA).

@@

```{r echo=FALSE,fig.cap="**Figure 6. Mathematical modelling and signalling mechanisms of the cPRC circuit.** (**A**) Diagram of the mathematical model with componenets, interactions, parameters and equations used to model Ca^2+^ dynamics. (**B**) Average Ca^2+^ traces (upper) and traces from model simulations (lower) fitted to the data. (**C**) Dot plot of genes (columns) expressed in three types of cells (rows) in the cPRC circuit using single cell RNA-Seq. The size of the dots is expressed in proportion to the percentage of cells expressing that gene relative to all cells. The colours represent the normal logarithm of the number of transcripts in the cells expressing the gene. (**D**) Schematic diagram of the signalling pathway of the cPRC circuit, focusing on the NO feedback. "}

@@

```{r echo=FALSE,fig.cap="**Figure 7. Magnitude of cPRC activation as  a function of stimulus duration and intensity.** Median values of the $AUP$ are presented as a heatmap with dark blue indicating flat $C_P(t)$ trajectories with median $AUP$ close to 0 and red indicating that the $C_P(t)$ trajectories have a clearly defined delayed activation peak. "}

@@

NO is a free radical with a millisecond-to-second half life and thus a limited signalling range.
In neuronal signalling, NOS is often localized to neurites [@Kuntz2017] at close proximity to sGC at synapses [@burette2002; @garthwaite2015a].
In the cPRC circuit, NOS is localised to the dendritic compartment of INNOS cells where also cPRC postsynaptic sites occur.
NIT-GC1, the target of NO signalling is localised to cPRC projections.
NO-mediated retrograde signalling thus likely occurs in the neurosecretory plexus where NOS- and NIT-GC1-containing projections are in close proximity and where we detected NO production following UV stimulation.
In contrast, INNOS to INRGWa synapses occur outside the neurosecretory plexus in the ventral brain neuropil.

@@

NIT-GC2-dependent signalling is required for the feedforward activation of the INNOS cells through an unknown transmitter.
The activation of INNOS leads to NOS activation and NO release, potentially through a canonical Ca^2+^-calmodulin pathway.
Our model predicts feedforward inhibiton from INNOS to INRGWa, possibly mediated by glycin transmission.
NO released by INNOS neurites activates NIT-GC1 in cPRC projections that could lead to cGMP production and the opening of CNGAα.
This late-phase cPRC activation is not directly dependent on the UV/violet signal and can happen post-stimulus.
The late-phase cPRC activation leads to INRGWa activation via an unknown transmitter that is likely different from the one acting on INNOS.

In addition, the cPRC circuit expresses several neuropeptides and their receptors, suggesting further neuromodulatory signals.
For example, INRGWa express the proneuropeptide RGWamide and cPRCs and INNOS express its receptor, suggesting retrograde peptidergic signalling in the circuit.

@@

We hypothesise that the differential beating of prototroch versus trunk cilia causes a head-up or head-down orientation due to physics alone.
During the UV response, prototroch cilia beat slower than trunk cilia, resulting in a head-down stable state ('rear-wheel drive').
In contrast, during the pressure response prototroch cilia beat faster than trunk cilia [@bezares-calderon2023], leading to a head-up orientation ('front-wheel drive').
Testing this hypothesis will require biophysical experiments and mathematical modelling.

@@

Animals evolved distinct photosensory systems coupled to non-overlapping circuits and guiding unique behavioural responses.
These sensory systems can employ different opsin molecules and be tuned to different wavelengths of light.
The avoidance of noxious UV/violet light is often mediated by extraocular photoreceptors and their circuits, distinct from the pigmented visual eyes.
These two types of systems co-exist in *Platynereis* where the pigmented eyes and eyespots guide phototaxis with a maximum sensitivity to cyan light (\~500 nm) [@Randel2014a; @Veraszto2018].
Planarians also have pigmented visual eyes mediating phototaxis to cyan light and peripheral extraocular photoreceptors mediating UV avoidance behaviour [@Shettigar2021].
*Drosophila* larvae have cerebral eyes called Bolwig's organs involved in phototaxis [@kane2013] and several types of extraocular UV-sensory cells that tile the body wall and mediate UV avoidance [@Imambocus2022].

@@

For the generation of a *NOS* knockout we used an sgRNA targeting the third exon of the *Platynereis dumerilii* *NOS* (target site: 5'-GGGCAATACTGGCTCCACTC-3').
We selected the site to avoid polymorphic sites in our laboratory culture.
The sgRNA was assembled from two annealed oligonucleotides (5'-TAGGGCAATACTGGCTCCACTC-3', 5'-AAACGAGTGGAGCCAGTATTGC-3') forming overhangs for cloning into the BsaI site of the plasmid pDR27456 [\@hwang2013](42250,%20Addgene), which contains next to the BsaI site a tracrRNA sequence.
We used this plasmid to PCR amplify DNA (primers: T7, 5'-AAAAGCACCGACTCGGTGCC-3') for synthesizing the sgRNA.
To purify the PCR product, we used the QIAquick PCR Purification Kit (Qiagen).
We then synthesized sgRNA with the MEGAshortscript Kit (Thermo Fisher Scientific) and purified it with the MEGAclear Kit (Thermo Fisher Scientific).
To produce the Cas9-mRNA, we used the plasmid (pUC57-T7-RPP2-Cas9) containing the Cas9 ORF fused to 169 base pair 5' UTR from the *Platynereis dumerilii* 60S acidic ribosomal protein P2.
After in vitro transcription with the mMessage mMachine Kit (Thermo Fisher Scientific), we capped and polyA-tailed the Cas9-mRNA with the Poly(A) Tailing Kit (Thermo Fisher Scientific).
We coinjected the sgRNA (18 ng/ml) and the Cas9-mRNA (180 ng/µl) into fertilized eggs of *Platynereis dumerilii* wild-type parents according to an established injection procedure [@Conzelmann2013a].
Eggs were kept at 18°C for 45 min before injection and were injected at 14.5°C.
The injected individuals were kept at 18°C for 5 to 8 days in 6-well-plates (Nunc multidish no. 150239, Thermo Scientific) and then cultured at 22°C until sexual maturity.
The mature worms were crossed to wild-type worms and the progeny was genotyped, resulting in two founder lines, which were bred to homozygosity.

@@

We assayed larval photoresponses in a vertical Plexiglas column (31 mm x 10 mm x 160 mm water height).
For stimulation, we illuminated the column from the top with a monochromator (Polychrome II, Till Photonics) controlled by AxioVision 4.8.2.0 (Carl Zeiss GmbH, Jena) via analog voltage.
The light passed a collimator lens (LAG-65.0-53.0-C with MgF2 Coating, CVI Melles Griot) before entering the column.
The column was illuminated from both sides with light-emitting diodes (LEDs).
The LEDs on each side were grouped into two strips.
One strip contained UV (395 nm) LEDs (SMB1W-395, Roithner Lasertechnik) and the other infrared (810 nm) LEDs (SMB1W-810NR-I, Roithner Lasertechnik).
The UV LEDs were run at 4 V. The infrared LEDs were run at 8 V (overvoltage) to illuminate the larvae for imaging with a DMK camera (DMK 22BUC03, The Imaging Source).
We recorded videos at 15 frames per second with the software IC Capture (The Imaging Source).

To compare the behavior of wildtype and NOS-knockout larvae at 3 days in the vertical column, the larvae were mixed and left in the dark for 5 min.
The larvae were treated with NOS inhibitors for pharmacology.
The NOS inhibitors were L-NAME.
The larvae were treated with different concentrations in adjacent columns.
The concentrations for the NOS inhibitors were control, 1 mM, 0.1 mM.
The larvae were recorded for 1 min in the dark followed by exposure to collimated cyan (480 nm) light from the top of the column for 2 min, then 2 min darkness, and finally collimated UV (395 nm) light from the top of the column for 2 min.
Stimulus light was provided by the monochromator (Polychrome II, Till Photonics).
Scripts are available at <https://github.com/JekelyLab/NOS>.

@@

To identify NOS sequences, we obtained a seed database of oxygenase domains from the Pfam database (PF02898).
From these sequences, we produced a Hidden Markov Model (HMM) and used this to mine the 47 metazoan, 2 choanoflagellate and 2 filasterea species investigated.
HMM models were run in HMMR3 with an e-value of 1e−15.
We ran CD-Hit [@Fu2012] to eliminate redundant sequences (at a 80% threshold).
We aligned the sequences with MAFFT version 7, with the iterative refinement method E-INS-i.
Alignments were trimmed with TrimAl in gappy-out mode [@Capella-Gutiérrez2009].
To calculate maximum-likelihood trees, we used IQ-tree2 with the LG+G4 model.
To calculate branch support, we ran 1000 replicates with the aLRT-SH-like and aBayes methods [@Minh2020].
The sequences used for the phylogenetic analysis are available in **Supplementary file 1**, the trimmed alignment is available in **Supplementary File 2** and the pre-trimmed data in **Supplementary file 3**.

@@

To identify NIT-GCs, we obtained a seed database of adenylate and guanylate cyclase catalytic domains from the Pfam database (PF00211).
From these sequences, we produced a Hidden Markov Model (HMM) and used this to mine the 45 metazoan, 2 choanoflagellate and 2 filasterea species investigated.
HMM models were run in HMMR3 [@Eddy2011] with an e-value of 1e−15.
We ran CD-Hit [@Fu2012] to eliminate redundant sequences (at a 80% threshold).
To identify clusters, we used sequence-similarity-based clustering in Clans [@frickey2004] and the convex-clustering option with 100 jack-knife replicates.
The NIT-GCs are extremely well conserved within the membrane-bound guanylate cyclases and form an easily recognizable cluster.
To analyze the phylogeny of NIT-GCs, the cluster containing these GCs together with membrane-bound guanylate cyclases were parsed and used for tree building.
We aligned the sequences with MAFFT version 7, with the iterative refinement method E-INS-i[@Katoh2013].
Alignments were trimmed with TrimAl in gappy-out mode [@Capella-Gutiérrez2009].
To calculate maximum-likelihood trees, we used IQ-tree2 with the LG+G4 model.
To calculate branch support, we ran 1000 replicates with the aLRT-SH-like and aBayes methods [@Minh2020].
The sequences used for the phylogenetic analysis are available in **Supplementary File 4**, the trimmed alignment is available in **Supplementary File 5** and the pre-trimmed data in **Supplementary file 6**.

@@

Whole-mount immunostaining of *Platynereis* larvae fixed with 4% paraformaldehyde was carried out using the primary antibodies listed in KEY RESOURCES TABLE. Immunostainings were carried out as previously described [@Conzelmann2012].

>>>>

Nitric oxide (NO) is one such modulator with unique physical and signalling properties.
This free radical synthesized from L-arginine by nitric oxide synthase (NOS), is short-lived and can diffuse across biological membranes [@Cudeiro1999; @thomas2015].
Canonical NO signalling involves the Ca^2+^/calmodulin-dependent activation of NOS, NO production and diffusion, and the NO-dependent activation of soluble guanylate cyclases (sGC) leading to cGMP production [@Bredt1990; @Hoelscher1997].
Given that NOS activation is Ca^2+^ dependent and NOS shows neuron-type-specific expression [@Gibbs1998a; @Wildemann1999; @Aso2019; @Mobley2022], NO action can lead to the activity-dependent modulation of neural circuits at specific sites [@Wang2007; @Vielma2014; @jacoby2018; @Aso2019] .

In the vertebrate retina, NOS is expressed in amacrine, ganglion and other cells and the actions of NO can be diverse [@Cudeiro1999; @Wang2007; @jacoby2018].
For example, defective NO signalling in NOS knockout mice leads to a decreased sensitivity of retinal ganglion cells to light stimulation [@Wang2007].
In the retina, NO signalling can also involve pathways other than canonical sGC signalling [@Tooker2013; @Wei2012; @jacoby2018].
Due to the complex expression of NOS in vertebrates and the diversity of its functions, it has been challenging to link the neurophysiological effects of NO to signalling mechanisms and behaviour change.

@@

```{r echo=FALSE,fig.cap="**Figure 2. NO produced by UV/violet stimulation to cPRCs.** (**A**) DAF-FM fluorescence in the region of the neurosecretory neuropil. The white line indicates the outline of the larva. The dashed line corresponds to the area where fluorescence was quantified. The circles indicate the location of cPRC and control stimulation. The cPRCs are marked by thin lines. (**B**) Changes in DAF-FM fluorescence before and during 405 nm light stimulation. (**C**) Changes in DAF-FM fluorescence over time during 405 nm stimulation of the cPRCs (green) or a control area (ctr: gray). The purple box indicates the duration of 405 nm stimulation. Individual traces normalized (ΔF/F0) are shown as thin lines. Thick lines show the mean value with 0.95 confidence intervals. N = 9 larvae for control and 11 for cPRC stimulation."}

@@

To characterise the expression of NIT-GCs, we used previously published spatially mapped single-cell transcriptome data [@Achim2015; @williams2017].
Among the 15 NIT-GCs, two showed high and specific expression in the cPRCs and one was expressed in the INNOS cells (Figure 4---figure supplement 2).
In the single-cell data, we could identify the cPRCs by the specific expression of the markers *c-opsin1* and the *pedal-peptide2 neuropeptide precursor* (*MLD proneuropeptide*) [@Arendt2004; @williams2017] (Figure 4---figure supplement 3A).
The INNOS cells were identified by NOS expression and spatial mapping in the brain [@Achim2015].
We decided to focus on two NIT-GCs expressed in the cPRCs and with a full-length sequence, *NIT-GC1* and *NIT-GC2*.
To confirm the single-cell data, we first carried out in situ hybridisation chain reaction (HCR) with probes for *NIT-GC1* and *NIT-GC2* mRNA. Both genes were specifically expressed in the four cPRCs, as confirmed by co-labeling with an acetylated α-tubulin antibody and with an HCR probe against *pedal peptide 2/MLD proneuropeptide* (Figure 4C, D and Figure 4---figure supplement 3A-C).
To analyse the subcellular localisation of NIT-GC1 and NIT-GC2 at the protein level, we raised and affinity-purified polyclonal antibodies against a specific peptide sequence from both proteins.
In immunostainings, we found that NIT-GC1 was localise to the region corresponding to the axonal projections of the cPRCs in the anterior neurosecretory plexus (Figure 4E).
Co-immunostaining with the rabbit NIT-GC1 antibody and a custom rat antibody raised against *Platynereis* NOS revealed the localisation of both proteins in dots in the neurosecretory neuropil (Figure 4---figure supplement 3D).
In contrast, NIT-GC2 specifically labelled the ramified sensory cilia of the cPRCs (Figure 4F).
These different subcellular localisations suggest that the two NIT-GCs are involved in different intracellular signalling processes in the ciliary and axonal regions of the cPRCs.

@@

We first quantified the responses of the INNOS and INRGWa interneurons during 405 nm stimulation of the cPRCs.
In both wild type and *NOS*-mutant larvae, INNOS cells showed an increase in Ca^2+^ during stimulation (Figure 5D).
In contrast, the INNOS response was flat or slightly negative in NIT-GC2 morphant larvae (Figure 5E) revealing an essential role for NIT-GC2-mediated cPRC suppression in INNOS activation.
The INRGWa cells were initially inhibited during cPRC stimulation, followed by a delayed activation paralleling the second activation phase of the cPRCs.
This late INRGWa response was lacking in *NOS*-mutants (Figure 5F).
In NIT-GC2 morphants, INRGWa cells showed a transient increase in Ca^2+^ that decayed after light off and a delayed activation was not present (Figure 5G).

Next, we imaged Ca^2+^ signals from Ser-h1 and MC neurons in wild type and *NOS* mutant larvae.
Ser-h1 cells showed an activation profile that correlated with cPRC activity, including a reduction in Ca^2+^ during stimulation followed by a rebound, a response that was defective in *NOS* mutants (Figure 5I).
The MC cell showed sustained activation, including a late-phase that was lacking in *NOS* mutants (Figure 5I).
These data suggest that during 405 nm stimulation the Ser-h1 cells are inhibited and the MC cell is activated, and this regulation is NO-dependent.
This pattern is expected to inhibit ciliary activity in the prototroch but not in the other ciliary bands (with no Ser-h1 or MC synapses), triggering NO-dependent downward swimming.

@@

To further analyse the dynamics of responses to UV light and formally describe cPRC phototransduction, synaptic connections and NO retrograde signalling, we developed a mixed cellular-circuit-level mathematical model.
We used our Ca^2+^ imagining data of cPRC, INNOS and INRGWa cells collected in wild type, *NOS* knockout and *NIT-GC2* morphant larvae to formulate assumptions that are the basis of the proposed model.

We modeled a direct c-opsin1-dependent response to UV leading to the initial rise in Ca^2+^ levels.
The mechanism of this signal is not known but may include Gα and Gβ signalling, CNGAα channel activation or other pathways.
We further assumed that in cPRC cells, Ca^2+^ levels increase with cGMP.
We modeled a decrease in cGMP and Ca^2+^ levels due to a NIT-GC2-dependent pathway (Figure 6A).
The effect of NO in cPRC cells is captured by another term that describes a NIT-GC1 and NO-dependent increase in cGMP.
Based on the synaptic connectome, we infered a feedforward coupling between cPRC cells and INNOS and INRGWa cells.
The proposed synaptic coupling is UV-dependent, decays linearly and is suppressed in NIT-GC2 morphants (Figure 6A).

In the INNOS cells, we assumed that excitatory synaptic input from cPRC (or rebound from tonic inhibition) leads to a rise in Ca^2+^ leading to NO production via NOS.
This input is NIT-GC2-dependent.
In the INRGWa cells, we modeled a decrease in Ca^2+^ based on an inhibitory UV-dependent synaptic signal from the cPRCs.
We further assumed the existence of direct excitatory coupling between cPRC Ca^2+^ levels and INRGWa Ca^2+^ to account for the late effects of cPRC on INRGWa (Figure 6A).
The UV-dependent and Ca^2+^-dependent signals may be mediated by different transmitters released by cPRCs during distinct phases of their activation cycle.
In addition, we assumed a direct inhibitory coupling between INNOS Ca^2+^ and INRGWa Ca^2+^.
Finally, in all variables, we assumed a linear decay and constant production to set a steady state (Figure 6A).
Since the aim of the model is to capture the normalised fluorescence data, the model is nondimensionalised.
UV stimulation is modelled as a square pulse.
To find model parameters producing an output fitting the experimental data we employed a global optimisation method known as a genetic algorithm (GA).

@@

```{r echo=FALSE,fig.cap="**Figure 6. Mathematical modelling and signalling mechanisms of the cPRC circuit.** (**A**) Diagram of the mathematical model with the componenets, interactions, parameters and equations used to model Ca^2+^ dynamics. (**B**) Average Ca^2+^ traces (upper) and traces from model simulations (lower) fitted to the data. (**C**) Dot plot of genes (columns) expressed in three types of cells (rows) in the cPRC circuit using single cell RNA-Seq. The size of the dots is expressed in proportion to the percentage of cells expressing that gene relative to all cells. The colours represent the normal logarithm of the number of transcripts in the cells expressing the gene. (**D**) Schematic diagram of the signalling pathway of the cPRC circuit, focusing on the NO feedback. <br> Figure 6 -- source data 1. TPM values for each gene and the percentage of expressed genes."}

@@

```{r echo=FALSE,fig.cap="**Figure 7. Magnitude of cPRC activation as  a function of stimulus duration and intensity.** Median values of the $AUP$ are presented as a heatmap with dark blue indicating flat $C_P(t)$ traces with median $AUP$ close to 0 and red indicating that the $C_P(t)$ traces have a clearly defined delayed activation peak. "}

@@

NO is a free radical with a millisecond-to-second half life and thus a limited signalling range.
In neuronal signalling, NOS is often localised to neurites [@Kuntz2017] close to sGC at synapses [@burette2002; @garthwaite2015a].
In the cPRC circuit, NOS is localised to the dendritic compartment of INNOS cells where also cPRC postsynaptic sites occur.
NIT-GC1, the target of NO signalling is localised to cPRC projections.
NO-mediated retrograde signalling thus likely occurs in the neurosecretory plexus where NOS- and NIT-GC1-containing projections are in close proximity and where we detected NO production after UV stimulation.
In contrast, INNOS to INRGWa synapses occur outside the neurosecretory plexus in the ventral brain neuropil.

@@

NIT-GC2-dependent signalling is required for the feedforward activation of the INNOS cells through an unknown transmitter.
The activation of INNOS leads to NOS activation and NO release, potentially through a canonical Ca^2+^-calmodulin pathway.
Our model predicts feedforward inhibition from INNOS to INRGWa, possibly mediated by glycin transmission.
NO released by INNOS neurites activates NIT-GC1 in cPRC projections that could lead to cGMP production and the opening of CNGAα.
This late-phase cPRC activation does not directly depend on the UV/violet signal and can happen post-stimulus.
The late-phase cPRC activation leads to INRGWa activation via an unknown transmitter that is likely different from the one acting on INNOS.

In addition, the cPRC circuit expresses several neuropeptides and their receptors, suggesting further neuromodulatory signals.
For example, the INRGWa cells express the proneuropeptide RGWamide and cPRCs and INNOS express its receptor, suggesting retrograde peptidergic signalling in the circuit.

@@

We hypothesize that the differential beating of prototroch versus trunk cilia causes a head-up or head-down orientation due to physics alone.
During the UV response, prototroch cilia beat slower than trunk cilia, resulting in a head-down stable state ('rear-wheel drive').
In contrast, during the pressure response prototroch cilia beat faster than trunk cilia [@bezares-calderon2023], leading to a head-up orientation ('front-wheel drive').
Testing this hypothesis will require biophysical experiments and mathematical modelling.

@@

Animals evolved distinct photosensory systems coupled to non-overlapping circuits and guiding unique behavioural responses.
These sensory systems can employ different opsin molecules and be tuned to different wavelengths of light.
The avoidance of noxious UV/violet light is often mediated by extraocular photoreceptors and their circuits, distinct from the pigmented visual eyes.
These two types of systems co-exist in *Platynereis* where the pigmented eyes and eyespots guide phototaxis with a maximum sensitivity to cyan light (\~500 nm) [@Randel2014a; @Gühmann2015; @Veraszto2018].
Planarians also have pigmented visual eyes mediating phototaxis to cyan light and peripheral extraocular photoreceptors mediating UV avoidance behaviour [@Shettigar2021].
*Drosophila* larvae have cerebral eyes called Bolwig's organs involved in phototaxis [@kane2013] and several types of extraocular UV-sensory cells that tile the body wall and mediate UV avoidance [@Imambocus2022].

@@

For the generation of a *NOS* knockout, we used an sgRNA targeting the third exon of the *Platynereis dumerilii* *NOS* gene (target site: 5'-GGGCAATACTGGCTCCACTC-3').
We selected the site to avoid polymorphic sites in our laboratory culture.
The sgRNA was assembled from two annealed oligonucleotides (5'-TAGGGCAATACTGGCTCCACTC-3', 5'-AAACGAGTGGAGCCAGTATTGC-3') forming overhangs for cloning into the BsaI site of the plasmid pDR27456 [\@hwang2013](42250,%20Addgene), which contains next to the BsaI site a tracrRNA sequence.
We used this plasmid to PCR amplify DNA (primers: T7, 5'-AAAAGCACCGACTCGGTGCC-3') for synthesizing the sgRNA.
To purify the PCR product, we used the QIAquick PCR Purification Kit (Qiagen).
We then synthesized sgRNA with the MEGAshortscript Kit (Thermo Fisher Scientific) and purified it with the MEGAclear Kit (Thermo Fisher Scientific).
To produce the Cas9-mRNA, we used the plasmid (pUC57-T7-RPP2-Cas9) containing the Cas9 ORF fused to 169 base pair 5' UTR from the *Platynereis dumerilii* 60S acidic ribosomal protein P2.
After in vitro transcription with the mMessage mMachine Kit (Thermo Fisher Scientific), we capped and polyA-tailed the Cas9-mRNA with the Poly(A) Tailing Kit (Thermo Fisher Scientific).
We coinjected the sgRNA (18 ng/ml) and the Cas9-mRNA (180 ng/µl) into fertilized eggs of *Platynereis dumerilii* wild-type parents according to an established injection procedure [@Conzelmann2013a].
Eggs were kept at 18°C for 45 min before injection and were injected at 14.5°C.
The injected individuals were kept at 18°C for 5 to 8 days in 6-well-plates (Nunc multidish no. 150239, Thermo Scientific) and then cultured at 22°C until sexual maturity.
The mature worms were crossed to wild-type worms and the progeny was genotyped, resulting in two founder lines, which were bred to homozygosity.

@@

We assayed larval photoresponses in a vertical Plexiglas column (31 mm x 10 mm x 160 mm water height).
For stimulation, we illuminated the column from above with a monochromator (Polychrome II, Till Photonics) controlled by AxioVision 4.8.2.0 (Carl Zeiss GmbH, Jena) via analog voltage.
The light passed a collimator lens (LAG-65.0-53.0-C with MgF2 Coating, CVI Melles Griot) before entering the column.
The column was illuminated from both sides with light-emitting diodes (LEDs).
The LEDs on each side were grouped into two strips.
One strip contained UV (395 nm) LEDs (SMB1W-395, Roithner Lasertechnik) and the other infrared (810 nm) LEDs (SMB1W-810NR-I, Roithner Lasertechnik).
The UV LEDs were run at 4 V. The infrared LEDs were run at 8 V (overvoltage) to illuminate the larvae for imaging with a DMK camera (DMK 22BUC03, The Imaging Source).
We recorded videos at 15 frames per second with the software IC Capture (The Imaging Source).

We compared the behavior of wildtype and *NOS*-knockout larvae in the vertical column.
In the inhibitor experiments we compared wildtype larvae before and after treatment with 0.1 mM or 1.0 mM NOS inhibitor L-NAME (in seawater).
All larvae were three days oldt.
Before the experiment, the larvae were mixed and left in the dark for 5 min.
The larvae were recorded for 1.5 min in the dark followed by exposure to diffuse UV light (395 nm) from the side for 2 min.
Then the larvae were left for another 2 min in darkness followed by exposue to collimated cyan (480 nm) light from above for 2 min, then 2 min darkness, and finally collimated UV (395 nm) light from above for 2 min.
Scripts are available at <https://github.com/JekelyLab/Jokura_et_al_NOS>.

@@

To identify NOS sequences, we obtained a seed database of oxygenase domains from the Pfam database (PF02898).
From these sequences, we generated a Hidden Markov Model (HMM) to mine the genomes and transcriptomes of 47 metazoan, 2 choanoflagellate and 2 filasterea species.
The HMM model was run in HMMR3 with an e-value of 1e−15.
We ran CD-Hit [@Fu2012] to eliminate redundant sequences (at a 80% threshold).
We aligned the sequences with MAFFT version 7, with the iterative refinement method E-INS-i.
The alignment was trimmed with TrimAl in gappy-out mode [@Capella-Gutierrez2009].
To calculate a maximum-likelihood tree, we used IQ-tree2 with the LG+G4 model.
To calculate branch support, we ran 1000 replicates with the aLRT-SH-like [@Guindon2010] and aBayes methods [@Minh2020].

@@

To identify NIT-GCs, we obtained a seed database of adenylate and guanylate cyclase catalytic domains from the Pfam database (PF00211).
From these sequences, we generated a Hidden Markov Model (HMM) to mine the 45 metazoan, 2 choanoflagellate and 2 filasterea genomes and transcriptomes.
The HMM model was run in HMMR3 [@Eddy2011] with an e-value of 1e−15.
We ran CD-Hit [@Fu2012] to eliminate redundant sequences (at a 80% threshold).
To identify clusters, we used sequence-similarity-based clustering in Clans [@frickey2004] and the convex-clustering option with 100 jack-knife replicates.
The NIT-GCs are extremely well conserved within the membrane-bound guanylate cyclases and form an easily recognizable cluster.
To analyze the phylogeny of NIT-GCs, the cluster containing these GCs together with membrane-bound guanylate cyclases were parsed and used for tree building.
We aligned the sequences with MAFFT version 7, with the iterative refinement method E-INS-i[@Katoh2013].
The alignment was trimmed with TrimAl in gappy-out mode [@Capella-Gutierrez2009].
To calculate a maximum-likelihood tree, we used IQ-tree2 with the LG+G4 model.
To calculate branch support, we ran 1000 replicates with the aLRT-SH-like and aBayes methods [@Minh2020].

@@

Whole-mount immunostaining of *Platynereis* larvae fixed with 4% paraformaldehyde was carried out using the primary antibodies listed in the KEY RESOURCES TABLE. Immunostainings were carried out as previously described [@Conzelmann2012].
